---
title: "bedscout"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{bedscout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bedscout)
```

# Loci enrichment analysis with bedscout

`bedscout` is a library that wraps some useful functionality around `BED` file
annotations mostly in the shape of `GRanges` objects. Currently implemented 
functions belong to one of the following categories:

* **Arithmetic**: Overlap-related simple operations that can be performed
between two or more `GRanges` objects, currently: `loci_overlap`, `bp_overlap`, 
`fc_enrichment` and `jaccard_index`, `loci_consensus`.

* **Pairwise**: Scoring functions to compare lists of `GRanges` in a pairwise
all-to-all kind of manner. For these, `pairwise_score` computes any arbitrary
score function (anything that takes two `GRanges` objects as inputs), including
the ones mentioned above in the **arithmetic** section. This returns a data
frame in long format (triplets gr1, gr2, score) that plays nice with plotting
libraries like `ggplot2`. A fixed example of this is `pairwise_enrichment`, 
which does exactly a `pairwise_score` with the scoring function fixed to 
`fc_enrichment`.

* **Annotate**: Intersect a `GRanges` object with corresponding named elements
from another `GRanges`. This can be gene TSS, ChromHMM segmentations and anything
really with a name field. These annotations can be done by proximity 
(`annotate_nearby_feature`), overlap (`annotate_overlapping_features`), or 
overlap maximizing jaccard index (`impute_feature`), especially useful for 
genomic segmentations like ChromHMM.

* **Venn**: Venn-diagram like operations. Right now, the only relevant function
is `calculate_venn_intersections` which calculates the intersections of all 
possible subsets in a `GRanges` list. This can be used to plot venn diagrams.

* **Plot**: Some handy plotting functions - `plot_euler` makes use of `eulerr`
package to make an Euler diagram of overlaps of a list of `GRanges` objects. 
`plot_pairwise_score` and `plot_combinations_score` are matched to the pairwise
operations and plot a heatmap with the specified score.

## Imputing elements from a genomic segmentation annotation

In many cases, one could have a set of genomic loci and would like to know what
type of genomic features they overlap with. 

`impute_features` takes a `GRanges` object and a named list of `GRanges`
objects, one per interest feature type. One classic example of this is
annotating genomic segmentations like ChromHMM. In those cases, it is almost
always the case that each loci in our interest group overlaps with more than
one feature, since genomic segmentations are partitions of the genome. 
`impute_features` takes all the overlap and select what is estimated as best
match, which is the one with maximal `jaccard_index`. This prevents flanking
empty segments (usually tagged as "quiescent") from being preferred, as overlap
tends to be significantly larger.

```{r}
genome_segment_file <- system.file(
  "extdata",
  "segments.bed",
  package = "bedscout"
)
``` 

`segments.bed` contains an example of the type of BED file, the `name` file
must be a category (i.e. a gene annotation file where each locus has a different
gene name would not be useful to process this way).

```{r}
cat(readLines(genome_segment_file), sep = '\n')
```

for instance:

```{r}
gr <- GenomicRanges::GRanges(
  seqnames = c("chr1"),
  IRanges::IRanges(4971, 5050),
  strand = "+"
)
```

Here we have a unique loci that completely contains a Type_C element but flanks
also some overlap with larger Type_A loci. We can ask `bedscout` to impute
the overlap that optimizes `jaccard_index`.

```{r}
# rtracklayer can import BED files into GRanges to use with impute_feature
gr_features <- rtracklayer::import(genome_segment_file)
gr_imputed <- impute_feature(gr, gr_features, "name")
gr_imputed
```

## Annotating features

Instead of imputing genomic segments, one might be interested in nearby genes,
for example. One can annotate those by using `annotate_nearby_features`. In this
case, all features within a given distance will be annotated, their names will
be comma-separated.

```{r}
peaks <- system.file(
  "extdata",
  "peaks_up.bed",
  package = "bedscout"
)

genes <- system.file(
  "extdata",
  "genes.bed",
  package = "bedscout"
)

peaks_gr <- rtracklayer::import(peaks)
genes_gr <- rtracklayer::import(genes)
annotate_nearby_features(
  peaks_gr,
  genes_gr,
  "name",
  distance_cutoff = 100,
  ignore.strand = FALSE
)
```

Note how, in the case of annotating genes, strand is probably relevant, whereas
in the case of genomic segments, it can be ignored. Generally, `bedscout` will
ignore strand unless stated otherwise, since there are many types of features 
that are not strand-specific and their corresponding bed-files might not even
have the field.

## Pairwise scoring

We could want to get a summary enrichment score of our `GRanges` object in each
of the categories in our segments.bed instead. `bedscout` has a generic 
`pairwise_score` function that takes two named lists of `GRanges` and provides
all pairwise scoring values in a data.frame in a long format, that can be 
easily plotted with `ggplot2`, for example.

We can import `segments.bed` to a `GRanges` named list using
`import_named_bed_into_list`:

```{r}
my_feature_list <- import_named_bed_into_list(genome_segment_file)
```

Now, `my_feature_list` contains one `GRanges` object for each `Type_A` to 
`Type_F` elements:

```{r}
my_feature_list
```

Now, we might have two different groups of loci in another list:

```{r}
gr1 <- GenomicRanges::GRanges(
  seqnames = c("chr1"),
  IRanges::IRanges(4971, 5050),
  strand = "+"
)

gr2 <- GenomicRanges::GRanges(
  seqnames = c("chr1"),
  IRanges::IRanges(10000, 10100),
  strand = "-"
)

gr_list <- list("genes_up" = gr1, "genes_down" = gr2)
```

And then we can calculate the pairwise overlap:

```{r}
pairwise_score(
  gr_list,
  my_feature_list,
  score_func = loci_overlap
)
```

This can be manually plotted, or make some default plot by using 
`plot_pairwise_score` function also included:

```{r}
plot_pairwise_score(
  gr_list,
  my_feature_list,
  score_func = jaccard_index
)
```

`score_func` can be any function that takes two `GRanges` objects and returns
a numerical value. 

## Combinations within the same list

It could also be possible to check possible pairwise combinations between a 
single list of `GRanges` elements, ignoring symmetrical calculations, provided
that we use a score function that is commutative. If a non-commutative operation
is specified, it cannot be guaranteed that the combinations will have a specific
order.

To illustrate this example, maybe you have three sets of peaks and want to
know how they overlap:

```{r}
peak_files <- system.file(
  "extdata",
  package = "bedscout"
) |> list.files(pattern = "peaks", full.names = TRUE)

peak_grlist <- lapply(peak_files, rtracklayer::import)

# Name your files:
names(peak_grlist) <- basename(peak_files)
```

Now you can plot a pairwise enrichment, maybe fixing the function
`fc_enrichment` with your genome size:

```{r}
gsize = 200000
enrichment_func <- purrr::partial(fc_enrichment, genome_size = gsize)

plot_combinations_score(
  peak_grlist,
  score = enrichment_func
)
```

## Euler diagrams

In cases where the amount of sets is not so large, one could want to check
overlaps in the form of an Euler diagram. This is done via the `eulerr` package,
and the intersections calculated by `bedscout` making use of `GRanges`
operations.

```{r}
plot_euler(peak_grlist, names = basename(peak_files))
```

It is possible to set the shape of the circles to be ellipses, to facilitate
difficult fits:

```{r}
plot_euler(peak_grlist, names = basename(peak_files), shape = "ellipse")

```

As you can see in the figure, the "1" intersection has a more similar size
to half of the "2", which is what we want. Note that some combinations might
not be possible to draw, and keep and eye for possible inconsistencies. 

The wonderful `eulerr` package has built-in diagnostics functions,
like error_plot. To check if your euler plot is accurately drawn, you can 
call `fit_values` instead and use the `error_plot` function to check it.

```{r}
fit_values <- fit_euler(peak_grlist, names = basename(peak_files), shape = "circle")
eulerr::error_plot(fit_values)
```

The size that "1" intersection is underrepresented. Check out
`eulerr` package [on CRAN](https://cran.r-project.org/web/packages/eulerr/index.html)
for more details about how these are calculated.


## Obtaining a consensus loci set out of a list of GRanges objects

Sometimes you might want to obtain a genomic annotation that is a consensus of 
a group of annotations. The typical example of this is peaks called over a set
of replicates. For that, you can use the `loci_consensus` function, which takes
a list of GRanges objects and returns a single GRanges object with the consensus
loci. The parameters to the function are `min_consensus`, which defaults to 1.
This means that any locus that appears in at least one of the lists will be
considered, so it is equivalent to merging all the ranges.

```{r}
loci_consensus(peak_grlist, min_consensus = 3)
```

It is also possible to resize the ranges to a fixed size before getting the 
consensus list:

```{r}
loci_consensus(peak_grlist, min_consensus = 2, resize = 500, anchor = "start")
```

